---
title: "Chapter 2 - Statistical Learning Notes"
author: "William Surles"
date: "April 12, 2017"
output:
  html_document: default
  html_notebook:
    highlight: tango
    self_contained: yes
    theme: flatly
---

***
> This document was updated on **`r Sys.Date()`**.

***

## Intro

  - Welp. Lets just get through all the code example.
  - I spend too long talking if I start so I'm just gonna do the code and I will
come back and add notes if I need to.
  - The book has the best notes. : )
  - Really, this doc is jut a reference for me to use later

```{r set-options, eval=T, echo=F, cache=F}

knitr::opts_chunk$set(eval=T, echo=T, cache=T, message=F, warning=F)

```


```{r}

library(stringr)

```


## 2.1 What is Statistical Learning?

## 2.2 Assessing Model Accuracy

## 2.3 Lab: Intro to R

### 2.3.1 Basic Commands

`c()` combines its inputs into a vector

```{r}

x <- c(1,3,2,5)
x

```

The `=` symbol works too

```{r}

y = c(1,4,3)
y

```

Use `?` to get help

``` {r, message=T}

print(?c)

```

The '+' function will add arrays by index (they must have the same length)

```{r}

x <- c(1,6,2)
y <- c(1,4,3)
x + y

```

```{r}

```


```{r}

```

```{r}

```

```{r}

```

```{r}

```

