---
title: "Chapter 2 - Statistical Learning Notes"
author: "William Surles"
output:
  html_document: default
  html_notebook:
    highlight: tango
    self_contained: yes
    theme: flatly
---

***
> This document was updated on **`r Sys.Date()`**.

***

## Intro

  - Welp. Lets just get through all the code examples.
  - I spend too long talking if I start so I'm just gonna do the code and I will
come back and add notes if I need to.
  - The book has the best notes. : )
  - Really, this doc is jut a reference for me to use later

```{r set-options, eval=T, echo=F, cache=F}

knitr::opts_chunk$set(eval=T, echo=T, cache=T, message=F, warning=F)

```


```{r}

library(stringr)
library(readr)
library(dplyr)
library(ggvis)
library(ISLR)

```


## 2.1 What is Statistical Learning?

## 2.2 Assessing Model Accuracy

## 2.3 Lab: Intro to R

***

### 2.3.1 Basic Commands

  - `c()` combines its inputs into a vector

```{r}

x <- c(1,3,2,5)
x

```

  - The `=` symbol works too

```{r}

y = c(1,4,3)
y

```

  - Use `?c` to get help with the `c()` or any function. 
  - The '+' function will add arrays by index (they must have the same length)
  - `?length` to see what the `length` function does. Or any function.

```{r}

x <- c(1,6,2)
y <- c(1,4,3)
length(x)
length(y)
x + y

```

  - `ls()` allows you to look at a list of all objects.
  - `rm()` is used to delete objects 
    - I rarely do this. 
  
```{r}
ls()
rm(x,y)
ls()

```

  - Remove all objects

```{r}
rm(list=ls())
```

  - `matrix()` will create a matrix
    - it needs 3 arguments, the data, the # of rows, and # of columns
    - by default it will create the matrix by columns, left to right

```{r}
x = matrix(data=c(1,2,3,4), nrow = 2, ncol = 2)
x
```

  - you don't need to give the names of the arguments
  - you can change all of the default arguments, for example build a matrix by row, rathter than by column
  - if you do not assign the matrix to a variable, it will just create it and print it to the screen
  
```{r}
x = matrix(c(1,2,3,4), 2, 2, byrow=T)
x
```

  - `sqrt()` will work on each element of the matrix
  - `x^2` will square each element of the matrix
  
```{r}
sqrt(x)
x^2
```

  - `rnorm` creates a vetor or random values sampled from the normal distribution
  - we get a different sample each time we call the function
  - we can change the values of the normal distribution. 
  - By default the mean is zero and sd is 1
  - `cor` will compute the correlation between two vectors
  
```{r}
x = rnorm(50)
y = x + rnorm(50, mean=50, sd = .1)
cor(x,y)
```

  - `set.seed` will allow you to create the exact same set of random numbers
  - this is great for teaching as it lets others get the same set of random numbers in the next functinos
  
```{r}
set.seed(1303)
rnorm(50)
```
  
  - `mean` calculates the mean of a vector of numbers
  - `var` calculates the variance of a vector of numbers
  
```{r}
set.seed(3)
y = rnorm(100)
mean(y)
var(y)

# calculate standard deviation
sqrt(var(y))
sd(y)
```

***

### 2.3.2 Graphics

  - I will probably use ggvis for all plots rather than base R. 
  - It is a big improvement on grammar and functionality
  - There may be some special R plots I may need to use occasionaly, but it seems to be rare.

```{r}
x = rnorm(100)
y = rnorm(100)
plot(x,y)
plot(x,y, 
     xlab = "this is the x-axis",
     ylab = "this is the y-axis",
     main = "Plot of X vs Y"
     )
```

  - similar plot but with ggvis
  
```{r}

data.frame(x=rnorm(100), y=rnorm(100)) %>% 
  ggvis(~x, ~y) %>%
  layer_points(fill := "blue")

```
  
  - you can save the output of an R plot 
  - you can use`pdf()` or `jpeg()`
  - `system` lets you can run a system command on the computer
    - here I make a separat folder to put images
    - `dev.off` says we are done creating the plot
  - these steps are useful to share the analysis results in an image in an automated way
    - once you have created an analysis and explained it, it may be useful to send updated results on a daily basis to a location. 
    - images are nice becasue they can easily be shared in an iframe or flow and be quickly viewed 
  
```{r}
system("mkdir images")
pdf("images/Figure.pdf")
plot(x,y,col="green")
dev.off()
```

  - `?seq` creates a sequence of numbers
  - There are many option to control length and step size
  
```{r}

x = seq(1:10)
x

## This is a shortcut
x = 1:10
x

x = seq(-pi, pi, length=20)
x

```

  - `contour` creates a contour plot of 3 dimensional data (like topographic map)
  - it takes an x, y, and z argument corresponding to x,y positions
  - there are many arguments to fine tune the plot
  
```{r}
y = x
f = outer(x, y, function(x,y) cos(y)/(1+x^2))
contour(x,y,f)

contour(x,y,f,nlevels=45, add=T)

fa = (f-t(f))/2
contour(x,y,fa, nlevels=15)

```

  - `image` works like contour, but it uses colors like a heatmap
  - `persp` creates a 3D plot. THeta and phi control the viewing angle
  
```{r}

image(x,y,fa)
persp(x,y,fa)
persp(x,y,fa, theta=30)
persp(x,y,fa, theta=30, phi=20)
persp(x,y,fa, theta=30, phi=70)
persp(x,y,fa, theta=30, phi=40)

```

***
### 2.3.3 Indexing Data
  - The book just covers matrices
  - It works similar for vectors, and dataframes.
  - I will mostly be working with dataframes so I added some exampoles here

```{r}
A = matrix(1:16,4,4)
A
A[2,3]
A[c(1,3), c(2,4)]
A[1:3,2:4]
A[1:2,]
A[,1:2]
A[1,]
A[-c(1,3),]
dim(A)
```

  - All of these work on a data frame also
  - dataframes can also be indexed by column names which reads a little cleaner
  
```{r}
df <- data.frame(a=1:4, b=5:8, c=9:12, d=13:16)
df
df[2,3]
df[c(1,3), c(2,4)]
df[1:3,2:4]
df[1:2,]
df[1:2] ## it defaults to the columns if you do not put in the ,
df[,1:2]
df[1,]
df[-c(1,3),]
dim(df)

df$a
df[c('a','b')]
df[1:2, c('a','b')]
df[c(1,4), c('a','d')]

```

***
### 2.3.4 Loading Data

  - I will probably only use readr since it is a newer library that replaces a lot of the code in this section
  - Auto is already part of the ISLR library so I don't need to load it
  - `fix` opens a spreadsheet viewer but its really slow
    - This has been replaced by the variable viewer built into R studio
  - `read_csv` is the kef function. It has many arguments to customize loading the data, but its defaults do a great job and make life easy. In most cases it does exactly what you would want. 
  
```{r}

## This just needs to be run the first time
# system("mkdir data")
# write_csv(Auto, "data/Auto.csv")

df_auto <- read_csv("data/Auto.csv")
str(df_auto)
glimpse(df_auto)
head(df_auto)
names(df_auto)

```

***
### 2.3.5 Aditional Graphical and Numerical Summaries

```{r}

plot(df_auto$cylinders, df_auto$mpg)

```

  - I typically use `ggvis` library for plots or `rcharts` if I want javascript charts. 
    - The base r plots are often used in old books like this but its worth learning the improved charting libraries.
  - I also never use `attach` because dplyr and ggvis let me reference the column name directly

```{r}

df_auto %>%
  ggvis(~cylinders, ~mpg) %>%
  layer_points(
    fill := "green",
    fillOpacity := .4
    )

```

  - `varwidth=T` makes the box width based on the number of data points in the group I think
    - Thats one way to add more info to the chart but it does look a little weird. 
    
```{r}

df_auto$cylinders <- as.factor(df_auto$cylinders)

plot(df_auto$cylinders, df_auto$mpg)
plot(df_auto$cylinders, df_auto$mpg, col = 'red', varwidth=T, horizontal=T)
plot(df_auto$cylinders, df_auto$mpg, col = 'red', varwidth=T, xlab="cylinders", ylab = "MPG")

```

  - Its still a good idea to make the cylnders variable a categorical variable rather than numerical. i.e. make it a factor. 
    - This lets the boxplot skip the 7 which has no values
    
```{r}

df_auto %>%
  ggvis(~cylinders, ~mpg) %>%
  layer_boxplots(
    width = .9,
    fill := "green",
    fillOpacity := .4
    )

```


```{r}

hist(df_auto$mpg)
hist(df_auto$mpg, col=2)
hist(df_auto$mpg, col=2, breaks=15)

```


```{r}

attach(df_auto)
pairs(Auto)
pairs(~ mpg + displacement + horsepower + weight + acceleration)

```

  - I'm having trouble getting the `identify` function to work 
    - I think the best way to do this is by adding a tooltip to the ggvis chart
    
```{r}

plot(horsepower, mpg)
identify(horsepower, mpg, name)

```
  
  - You can create a tooltip for the ggvis chart pretty easily
  - You just need a function that describes what to print and how
  - then you call that function in the `add_tooltip` function. 
  - There rae plenty of examples to follow
  
```{r}

all_values <- function(x) {
  if(is.null(x)) return(NULL)
  str_c(names(x), ": ", format(x), collapse = "<br />")
}

df_auto %>%
  ggvis(~horsepower, ~mpg) %>%
  layer_points(
    fill := "green",
    fillOpacity := .6
  ) %>%
  add_tooltip(all_values, "hover")

```



```{r}


```



```{r}


```


```{r}


```


```{r}


```


```{r}


```


```{r}


```
